---
title: "Get Started with atlantio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Get Started with atlantio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(atlantio)
```


# Goal

The goal of atlantio is to easily manipulate Atlantis models [@audzijonyte_AtlantisSpatiallyExplicit_2019]. This means reading, editing and validating Atlantis inputs and outputs (atlantis + i/o = atlantio).

A major initial objective while working on the package was to be able to readily set up a semi-automatic calibration of the Lake Erie model using `calibrar` as described in @morell_ManualSemiautomated_2026.

This vignette is a quick tour of what the package currently does:

1. build an `Atlantis` object from input and output files,
2. read and write `.prm` parameter files,
3. extract time series and biomass tables from model outputs,
4. derive diet tables and food webs from the biology parameters,
5. list Atlantis parameters, generate template files and prepare a calibration.


# Example files

The package ships with a tiny Atlantis model (inputs and outputs of a short run). `atlantis_examples()` returns the path to these files, so every example below can be run as is.

```{r examples}
atlantis_examples("inputs") |> list.files()
atlantis_examples("outputs") |> list.files()
```


# Building a model

An `Atlantis` object is a container with one slot per kind of Atlantis file (geometry, run file, biology parameters, group file, main output, biomass tables, diet tables, etc.). `new_atlantis()` creates an empty one and printing it shows which components are available.

```{r empty}
mod <- new_atlantis()
mod
```

`atlantis_load_files()` reads files, detects their type from their extension and content, and stores them in the right slot. Here we add the geometry, the biology parameters, the group file and the run file.

```{r load-inputs}
mod <- mod |>
  atlantis_load_files(c(
    atlantis_examples("inputs", "tiny.bgm"),
    atlantis_examples("inputs", "tiny_biol.prm"),
    atlantis_examples("inputs", "tiny_groups.csv"),
    atlantis_examples("inputs", "tiny_run.prm")
  ))
mod
```

Each component is accessible with `@`. The geometry is an `sf` object, the group file a data frame, and parameter files are named lists.

```{r slots}
mod@geometry[, c("label", "botz", "area")]
mod@group[, c("Code", "Name", "GroupType", "NumCohorts", "IsPredator")]
mod@biology$mum_YPH
mod@run$tstop
```

Files can be added at any time, for instance outputs once the model has been run (see below).


# Reading and writing parameter files

`.prm` files are the main way to configure Atlantis. `read_prm()` parses them into a named list where every entry keeps its comments as an attribute, and `write_prm()` does the reverse. This makes it easy to edit a few parameters programmatically and write a new file.

```{r prm}
prm <- read_prm(atlantis_examples("inputs", "tiny_run.prm"))
length(prm)
prm$tstop

prm$tstop <- 730
write_prm(prm[c("tstop", "dt", "toutstart", "toutinc")])
```

Pass a path to `file` to write to disk instead of the console.


# Working with outputs

Let's add the main NetCDF output and two biomass tables from the example run.

```{r load-outputs, message = FALSE}
mod <- mod |>
  atlantis_load_files(c(
    atlantis_examples("outputs", "output.nc"),
    atlantis_examples("outputs", "outputBiomIndx.txt"),
    atlantis_examples("outputs", "outputBoxBiomass.txt")
  ))
mod
```

## Biomass tables

`get_biomass_by_type()` returns one of the biomass tables (`"biomass_all"`, `"biomass_box"`, `"biomass_age"` or `"biomass_age_annual"`).

```{r biomass}
mod |>
  get_biomass_by_type("biomass_all") |>
  dplyr::select(Time, WAE, YPH, GZS, RDG) |>
  head()
```

## Time series from the NetCDF output

`create_time_series()` extracts any variable of the main output as a tidy data frame, optionally restricted to a set of boxes and layers. Time is in seconds since the start of the run.

```{r ts}
names(mod@main_output$var) |> head(10)

mod |>
  create_time_series(variable = "Temp", box = 2, layer = 1) |>
  head()
```

`create_time_series_biomass()` gathers reserve and structural nitrogen, numbers and weight for an age-structured group, per box, layer and age class.

```{r ts-biomass}
ts_yph <- mod |> create_time_series_biomass(group = "YPH")
head(ts_yph)
```

```{r ts-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
ts_yph |>
  dplyr::summarise(Nums = sum(Nums), .by = c(time, age)) |>
  ggplot(aes(x = time / 86400, y = Nums, colour = age)) +
  geom_line() +
  labs(x = "Day", y = "Numbers", title = "Yellow Perch abundance per age class")
```


# Diet table and food web

The biology parameter file contains the prey availability matrix (`pPREY` parameters). `create_diet_table()` reshapes it into a prey-by-predator table.

```{r diet}
diet <- create_diet_table(mod)
dim(diet)
diet[1:6, 1:6]
```

`create_foodweb()` goes one step further and returns the binary food web as an `igraph` object, ready for network analysis or plotting.

```{r foodweb}
fw <- create_foodweb(mod)
fw
```

```{r foodweb-plot, fig.height = 6}
par(mar = c(0, 0, 0, 0))
plot(
  fw,
  layout = igraph::layout_in_circle(fw),
  vertex.size = 16,
  vertex.color = "#f4e0b7",
  vertex.frame.color = "grey40",
  vertex.label.color = "black",
  vertex.label.cex = 0.75,
  edge.color = "grey60",
  edge.arrow.size = 0.3
)
```


# Parameters, templates and calibration

## Listing Atlantis parameters

atlantio embeds a description of Atlantis parameters extracted from the source code. `list_atlantis_parameters()` returns metadata about the Atlantis version and a list of parameters with their description, type and where they are read.

```{r parameters}
prm_list <- list_atlantis_parameters()
prm_list$meta$atlantis_version
length(prm_list$parameters)
prm_list$parameters[[1]]
```

See the [parameter set vignette](parameters/index.html) for a searchable table.

## Generating template files

`generate_file()` uses these descriptions to generate a parameter file with recommended values and comments. Values that need to be provided by the user are marked with `<<TO_EDIT>>`. Only the run file is supported for now.

```{r generate}
run_file <- generate_file("run", file = file.path(tempdir(), "run.prm"))
readLines(run_file, n = 12)
```

## Preparing a calibration

`generate_calibration_table()` turns a short YAML specification into the table of parameters to calibrate expected by `calibrar`: one row per parameter and position, with bounds, the transformation applied and the file the parameter belongs to.

```{r calibration}
readLines(atlantis_examples("calibrate", "mum.yaml")) |> cat(sep = "\n")

mod |>
  generate_calibration_table(atlantis_examples("calibrate", "mum.yaml")) |>
  head()
```

`transform_parameter_value()` and `inverse_transform_parameter_value()` map values between the calibration scale and the scale used in the parameter file.

```{r transform}
transform_parameter_value(-3, "pow10")
inverse_transform_parameter_value(0.001, "pow10")
```

The full workflow (running Atlantis, computing the objective function and driving `calibrar`) is described in the [calibration vignette](calibration/index.html).


# References
